\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-11 paper.tex
Added Handwritten digit: 1 dataset

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

In this paper, we introduce a novel method to determine the most
effective NLDR technique and the best hyperparameter choice that
provides the most useful representation of high-D data. Our approach
involves dividing the high-dimensional dataset into two parts: a
training set for constructing the model and a test set for generating
predictive values and residuals. Our algorithm takes a 2D embedding data
as the input and generate a tour that displays the high-dimensional
wireframe to overlay the data. The flow chart of the proposed algorithm
is shown in Figure~\ref{fig-meth}. The algorithm consists of two main
phases: (1) generating the model in the 2D space and (2) lifting the
model into high-D space. The main steps of the algorithm are described
in detail in this section using UMAP 2D embedding of the S-curve
dataset. This dataset has seven dimensions, including four noise
dimensions that were added to the original 3D data.

@@

In order to reduce the computational complexity associated with
performing NLDR techniques to high-D data, and as a initial step of
noise reduction of high-D data, PCA \citep[\citet{article68},
\citet{article69}]{article67} is applied as a preprocessing step.
Subsequently, the identified principal components, representing
directions of maximum variance, are used to construct the model.

Then, after performing NLDR, to have regular hexagons, 2D embedding data
are scaled accordingly.

@@

@@

The prediction approach involves employing the K-nearest neighbors (KNN)
algorithm. In this method, which operates as an unsupervised
classification problem, the nearest high-D model point (averaged high-D
point) is identified for the new high-D data point. Once the nearest
high-D model point is identified, its coordinates are used as the
predicted 2D embeddings for the new high-D data point. This step is
particularly valuable due to the limitations of certain NLDR techniques,
like tSNE, which don't provide a straightforward method for predicting
2D embeddings. As a result, our approach presents a solution that can
generate predicted 2D embeddings, irrespective of the NLDR technique
employed, effectively addressing this gap in functionality.

@@

In the 2D representations created by all NLDR techniques, as shown in
\textbf{?@fig-nldervis5Gau}, except for PHATE, there are five distinct
clusters. In tSNE, these five clusters are closely located to each other
(see \textbf{?@fig-nldervis5Gau} (a)). However, in PHATE, two clusters
are closely positioned, while the other three are more distant (see
\textbf{?@fig-nldervis5Gau} (c)). In PaCMAP, one cluster is at the
center, and the remaining four are positioned in four different
directions (see \textbf{?@fig-nldervis5Gau} (e)). In TriMAP, two
clusters are close, although not as close as in PHATE, and the other
three are well-separated (see \textbf{?@fig-nldervis5Gau} (d)). In UMAP,
all clusters are arranged in a parallel manner, with three in one line
and the other two in a separate line (see \textbf{?@fig-nldervis5Gau}
(b)).

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of \textbf{?@fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of \textbf{?@fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of \textbf{?@fig-modelfiveGau} (c)). TriMAP and PaCMAP both display
five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of \textbf{?@fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures, as
indicated by lower AIC values in \textbf{?@fig-diagnosticpltGau} (a).

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-64-1.pdf}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

In this paper, we introduce a novel method to determine the most
effective NLDR technique and the best hyperparameter choice that
provides the most useful representation of high-D data. Our approach
involves dividing the high-D dataset into two parts: a training set for
constructing the model and a test set for generating predictive values
and residuals. Our algorithm takes a 2D embedding data as the input and
generate a tour that displays the high-D wireframe to overlay the data.
The flow chart of the proposed algorithm is shown in
Figure~\ref{fig-meth}. The algorithm consists of two main phases: (1)
generating the model in the 2D space and (2) lifting the model into
high-D space. The main steps of the algorithm are described in detail in
this section using UMAP 2D embedding of the S-curve dataset. This
dataset has seven dimensions, including four noise dimensions that were
added to the original 3D data.

@@

To reduce computational complexity when applying NLDR techniques to
high-D data and to reduce noise presence, PCA \citep[\citet{article68},
\citet{article69}]{article67} is used as a preprocessing step. PCA
involves identifying principal components that maximize variance. These
components are then used as the high-D data for the algorithm.

@@

First, we prepare the 2D embedding data to fit within the bounds
required for regular hexagonal binning. To achieve this, we implement
two key scaling steps. Scale the first 2D embedding component to range
between \(0\) and \(1\), ensuring that all data points fall within this
normalized interval. Secondly, we scale the second 2D embedding
component to range between \(0\) and \(y_{max}\) (see
Equation~\ref{eq-equation3}).

\begin{equation}\protect\hypertarget{eq-equation1}{}{
 ar = \frac{r_2}{r_1}
}\label{eq-equation1}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation2}{}{
 hr = \frac{hb}{wb}
}\label{eq-equation2}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 y_{max} = \text{ceiling}(\frac{ar}{hr}) * hr
}\label{eq-equation3}\end{equation}

@@

The prediction approach involves employing the K-nearest neighbors (KNN)
algorithm. In this method, which operates as an unsupervised
classification problem, the nearest high-D model point (averaged high-D
point) is identified for the new high-D data point. Once the nearest
high-D model point is identified, its coordinates are used as the
predicted 2D embedding for the new high-D data point. This step is
particularly valuable due to the limitations of certain NLDR techniques,
like tSNE, which don't provide a straightforward method for predicting
2D embedding. As a result, our approach presents a solution that can
generate predicted 2D embedding, irrespective of the NLDR technique
employed, effectively addressing this gap in functionality.

@@

In the 2D representations created by all NLDR techniques, as shown in
Figure~\ref{fig-nldervis5Gau}, except for PHATE, there are five distinct
clusters. In tSNE, these five clusters are closely located to each other
(see Figure~\ref{fig-nldervis5Gau} (a)). However, in PHATE, two clusters
are closely positioned, while the other three are more distant (see
Figure~\ref{fig-nldervis5Gau} (c)). In PaCMAP, one cluster is at the
center, and the remaining four are positioned in four different
directions (see Figure~\ref{fig-nldervis5Gau} (e)). In TriMAP, two
clusters are close, although not as close as in PHATE, and the other
three are well-separated (see Figure~\ref{fig-nldervis5Gau} (d)). In
UMAP, all clusters are arranged in a parallel manner, with three in one
line and the other two in a separate line (see
Figure~\ref{fig-nldervis5Gau} (b)).

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of \textbf{?@fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of \textbf{?@fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of \textbf{?@fig-modelfiveGau} (c)). TriMAP and PaCMAP both display
five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of \textbf{?@fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures, as
indicated by lower AIC values in Figure~\ref{fig-diagnosticpltGau} (a).

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Gau-1.pdf}

\caption{\label{fig-nldervis5Gau}2D layouts from different NLDR
techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP
(n\_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n\_inliers = 5,
n\_outliers = 4, n\_random = 3), and (e) PaCMAP (n\_neighbors = 10, init
= random, MN\_ratio = 0.9, FP\_ratio = 2). Is there a best
representation of the original data or are they all providing equivalent
information?}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltGau-1.pdf}

\caption{\label{fig-diagnosticpltGau}Goodness of fit statistics from
different NLDR techniques applied to training five spherical Gaussian
cluster dataset. What is the best NLDR technique to represent the
original data in 2D?}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-68-1.pdf}

\hypertarget{handwritten-digits}{%
\subsection{Handwritten digits}\label{handwritten-digits}}

\begin{itemize}
\tightlist
\item
  Digit 1 (7877 images only)
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltMNIST-1.pdf}

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-4.pdf}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisMNISTPACMAP-1.pdf}

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/CNXjWZMyh_M})}
